# AR models

One building block of statistical time series analysis is the *autoregressive (AR) model*. As the name implies, an AR model presumes that past values of the series can be used to directly predict current values using a regression-like structure:

$$Y_t = \beta_0 + \beta_1 Y_{t-1} + \beta_2 Y_{t-2} + \ldots + \varepsilon_t$$

I wrote the equation above using the notation of linear regression. We will be expressing the same concept using slightly different notation below.

## AR(1) process

Let's take the simplest case, where each value of the series is predicted only by the previous value. We would name this an *autoregressive model with order 1* or *AR(1) model*.

::: {.callout-note}
Let $\boldsymbol{Y}$ be a time series random variable observed at regular time periods $T = \{1, 2, \ldots, n\}$. Let $\boldsymbol{\omega}$ be a white noise process observed at the same time periods. If,

$$Y_t = \phi Y_{t-1} + \omega_t$$

For some $\phi \in \mathbb{R}$ and all $t \in T$, then we say that $\boldsymbol{Y}$ is an **autoregressive process with order 1**
:::

In theory, the exact properties of an AR(1) process depend on both the autoregressive parameter $\phi$ as well as the specific type of white noise process denoted by $\boldsymbol{\omega}$. In practice, we often assume $\boldsymbol{\omega}$ to be Gaussian white noise, leaving $\phi$ as the only input which needs estimation.

```{r}
#| code-fold: true
#| fig-cap: "Nine AR(1) processes with the same innovations"
par(mfrow=c(3,3),mar=c(3.1,3.1,3.1,1.1))
set.seed(0106)
w <- rnorm(100)
phivec <- c(-1.5,-1.0,-0.5,0,0.3,0.6,0.9,1.0,1.5)
Y <- t(rep(w[1],9))
for (i in 2:100){Y <- rbind(Y, t(rep(w[i],9)+phivec*Y[i-1,]))}
for (j in 1:9) plot(Y[,j],main=bquote(phi==.(phivec[j])),
                    type='l',xlab=NA,ylab=NA,sub=NA)
```

The plots above provide examples of some specific properties of AR(1) processes:

 * When $\phi \gt 1$ or $\phi \lt -1$, we say the process is **explosive.** After a short-to-moderate 'fuse', the innovations begin to cumulate exponentially, rapidly heading toward infinity.
 
 * When $\phi = 1$, we say that the process is a *random walk*, since we now have the familiar model $Y_t = Y_{t-1} + \omega_t$. When $\phi = -1$ we have a variant on a random walk in which both the odd and even observations begin closely-related random walks headed in opposite directions.
 
 * When $-1 \lt \phi \lt 1$ we observe a *stationary process*. This includes the case where $\phi = 0$ (pure white noise), but also other series with noticeable positive or negative autocorrelation. Despite the autocorrelation, the mean and variance of the unconditional variables $Y_t$ (that is, without knowing the past values of $\boldsymbol{Y}$) remain the same, and the autocorrelation of two values depends only on their lag distance, not their time index values.

## Generalizing to AR(*p*)

Now we may complicate the scenario, by allowing the current value $Y_t$ to have *several, different* relationships with *arbitrary* past lags of the series.

::: {.callout-note}
Let $\boldsymbol{Y}$ be a time series random variable observed at regular time periods $T = \{1, 2, \ldots, n\}$. Let $\boldsymbol{\omega} be a white noise process observed at the same time periods. If,

$$Y_t = \phi_1 Y_{t-1} + \phi_2 Y_{t-2} + \ldots + \phi_p Y_{t-p} + \omega_t$$

For some $(\phi_1, \phi_2, \ldots, \phi_p) \in \mathbb{R}^p$ and all $t \in T$, then we say that $\boldsymbol{Y}$ is an **autoregressive model with order** ***p***.
:::

These AR(*p*) processes can be difficult to visually distinguish because of the more complicated autocorrelation patterns.

```{r}
#| code-fold: true
#| fig-cap: "Nine AR(2) and AR(3) processes with the same innovations"
par(mfrow=c(3,3),mar=c(3.1,3.1,3.1,1.1))
set.seed(0107)
w <- rnorm(100)
phi1vec <- c(0.5,0.7, 0.7,0.6,0,  -1.6,0.5,0.3,0)
phi2vec <- c(0.5,0.7,-0.7,0.2,0.5, 0.8,0.4,0.2,0)
phi3vec <- c(0,  0,   0,  0,  0,   0,  0.3,0.1,1)
Y <- t(rep(w[1],9))
Y <- rbind(Y, t(rep(w[2],9)+phi1vec*Y[1,]))
Y <- rbind(Y, t(rep(w[3],9)+phi1vec*Y[2,]+phi2vec*Y[1,]))
for (i in 4:100){Y <- rbind(Y,
  t(rep(w[i],9)+phi1vec*Y[i-1,]+phi2vec*Y[i-2,]+phi3vec*Y[i-3,]))}
for (j in 1:9) plot(Y[,j],main=bquote(
  bold(phi) == list(.(phi1vec[j]),.(phi2vec[j]),.(phi3vec[j]))),
  type='l',xlab=NA,ylab=NA,sub=NA)
```

## Recognizing an AR process from its ACF plot

An autocorrelation function (ACF) plot will sometimes provide helpful summary or diagnostic information about an AR(*p*) model. In the simplest case of AR(1), the height of the ACF plot at the first lag will be equal to *p*, and the remaining lags will scale geometrically downward. For example, when $p=0.8$, the second lag will have an autocorrelation of roughly 0.64, and the third lag will have autocorrelation of roughly 0.512

```{r}
#| code-fold: true
#| fig-cap: "Two AR(1) processes and their ACF plots, 1000 obs each"
par(mfrow=c(1,2),mar=c(3.1,3.1,3.1,1.1))
set.seed(0108)
ar08 <- arima.sim(list(ar=0.8),n=1000)
arneg05 <- arima.sim(list(ar=-0.5),n=1000)

acf(ar08,main=expression(paste('ACF when ',phi == 0.8)),lag.max=15)
acf(arneg05,main=expression(paste('ACF when ',phi == -0.5)),lag.max=15)
```

More complex AR models have correspondingly complex ACF plots, and you will not always be able to easily diagnose the order and coefficient sizes:

```{r}
#| code-fold: true
#| fig-cap: "Two AR(2) processes and their ACF plots, 1000 obs each"
par(mfrow=c(1,2),mar=c(3.1,3.1,3.1,1.1))
set.seed(0109)
ar08neg05 <- arima.sim(list(ar=c(0.8,-0.5)),n=1000)
arneg05neg04 <- arima.sim(list(ar=c(-0.5,-0.4)),n=1000)

acf(ar08neg05,main=expression(paste('ACF when ',phi == list(0.8,-0.5))),lag.max=15)
acf(arneg05neg04,main=expression(paste('ACF when ',phi == list(-0.5,-0.4))),lag.max=15)
```

You do *not* need to be able to spot a complex AR model from a brief inspection of its time series or its ACF plot. However, you *should* know that complex behavior can be very accurately fit by a relatively simple AR model.

## Which AR processes are stationary

We will learn more about this elsewhere, when we discuss unit roots. For now, I will leave you with a simple set of guidelines:

 * AR(1) processes are stationary when $|\phi_1| \lt 1$
 
 * AR(2) processes are stationary when three conditions are met:
 
 $$\begin{aligned}(1) \qquad & |\phi_2| \lt 1 \\ (2) \qquad & \phi_1 + \phi_2 \lt 1 \\ (3) \qquad & \phi_1 - \phi_2 \lt 1 \end{aligned}$$
 
 Representing the two parameters $phi_1$ and $\phi_2$ on the coördinate plane, we would say any process with its parameters inside the shaded region is stationary:
 
```{r}
#| code-fold: true
#| fig-cap: 'Stationary sets of AR(2) parameters'
plot(c(-2.25,2.25),c(-1.25,1.25),type='n',main=NA,sub=NA,
     ylab=expression(phi[2]),xlab=expression(phi[1]),asp=1)
polygon(x=c(-2,0,2),y=c(-1,1,-1),lty=1,lwd=2,density=-1,col='#0000ff7f')
abline(h=0,v=0,lty=2,col='#7f7f7f')
```

 * For AR(3) and higher processes, stationarity will depend upon the roots of the characteristic polynomial, which can be calculated with a unit root test.
 
## AR model responses to system shocks

AR models translate one-time system shocks into persistent, slowly-decaying signals. Consider a simple AR(1) model which processes a fairly quiet series of innovations, interrupted irregularly by a much larger signal:

```{r}
#| code-fold: true
#| fig-cap: "Persistence and decay in an AR(1) model"
set.seed(0110)
w <- runif(100,-1,1)
w[8] <- 10; w[15] <- 10; w[25] <- -10; w[45] <- -10;
ar <- arima.sim(list(ar=0.9),n=100,innov=w)
plot(15+ar,type='s',lwd=2,ylim=c(-10,27),ylab=NA)
lines(w,type='h',col='#0000ff',lwd=2)
abline(h=15,lty=2,col='#7f7f7f')
legend(x='topright',lwd=2,col=c('#000000','#0000ff'),bty='n',
       legend=c(expression(paste('Time series ',Y[t])),
                expression(paste('Innovations ',omega[t]))))
```
